A Survey of Physical Design Techniques of Information Systems

### 2010-07-09 cs848-db/Paper.tex

B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but instead of having one value per node, a B-tee can have many values per node~
cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE). B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion. Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

@@

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the considerations taken for I/O operations is no longer there. The second is having index structures cache conscious. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{rao1999cache, boncz1999database}. This is because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive.

\subsubsection*{$B^{+}-Trees$}
The $B^{+}-Tree$, explained in the previous section, is originally designed for disk based database systems. However, it has also been found useful for main memory databases (REF) as it is cache conscious.

\subsubsection*{T Trees}
Lehmen and Carey~\cite{lehman1986study} study the different features needed for indices of MMDB. When designing data structures for MMDB, we are mainly concerned with the efficient use of CPU cycles and memory. Accordingly, the main goal of an index structure for MMDB is to minimize the overall computation time while using the minimum memory possible. Since all the database data is in main memory, index structures do not need to store actual values like those in disk-oriented databases. Instead, a pointer to the data instead of the actual data can be stored. Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the performance of the T Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures included were  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix of searches, inserts and deletes.

@@

\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
\cite{rao1999cache}.

\subsubsection*{Cache Sensitive T-Trees (CST-Trees)}

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned with the cache line size. The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache. Additionally, record pointers stored in the tree take up a lot of space.

To resolve these issues, Lee et al.~
cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary search tree in the array.

\subsubsection*{Cache Sensitive $B^{+}-trees$ ($CSB^{+}-trees)$}

Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main memory databases. However, traditional B+-Trees still had poor cache performance. Therefore, they propose Cache Sensitive B+-Trees ($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.

\subsubsection*{$Prefetching B^{+}-Trees (pB^{+}-trees)$}

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$. Since current computer systems can prefetch different data simultaneously, the author take advantage of this. The node size of the $B^{+}-tree$ is made wider than the cache line size. This way, several cache misses can be overlapped at the same time thus improving performance.

@@

>>>>

B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

@@

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

cite~\cite{dewitt1984implementation}.

\subsubsection*{B Trees}
Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B Trees in general have good cache behavior.

However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose Cache Sensitive B+-Trees
($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.

\subsubsection*{T Trees}
Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to
main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the
performance of the T
Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
was the first index structure specifically designed for main memory databases.

Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time~\cite{rao1999cache}. Although many keys are
fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem.

@@

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
Additionally, record pointers stored in the tree take up a lot of space.

To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
search tree whose node values are the maximum key of each node in the T Tre is constructed. That way, for a given value, the node containing this value could be
quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the necessary
nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each binary
search tree in the array.

\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.

\subsubsection*{Prefetching B+ Trees (pB+ Trees)}

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger size nodes means that we can overlap
multiple cache misses before they occur.

@@

Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
makes partitioning in memory unnecessary as it will not save any costs. However, partionining is needed in the secondary storage of the data on
disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
efficient in loading the important data such that the database users are not affected. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing}

\subsection{Clustering}
Clustering in disk based systems is important because sequential I/O access is cheaper than random or dispersed I/O access. This is not the case with main
memory, and so clustering is not needed in MMDB~\cite{garcia1992main, moldovan2008databases}. Therefore, components of one object may be spread across memory
without impacting performance.
